Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-17 2017-Xudong-SMT/FM-2018-extended.tex
Added the presentation of BIP, architectures and Failure Monitor, except for connectors

@@

@@

% Macros for the SOS rules and proof trees:
%\newcommand\openrule[2]{\redinfer{#1}{#2}}
\newcommand\openrule[2]{\inferrule*[myfraction=\reddottedrule,center]{#1}{#2}}
%\newcommand\openrule[2]{\inferrule*{#1}{#2}}
%\newcommand\ostate[1]{\triangleleft{\;#1\;}\triangleright}
\newcommand\ostate[1]{\triangleleft{#1}\triangleright}
\newcommand{\sm}[1]{\mbox{\boldmath\small #1}}
\usepackage{algorithm} 
%\usepackage{algorithmicx} 
\usepackage{algpseudocode}  
\renewcommand{\algorithmicrequire}{\textbf{Input:}} 
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\IfThen}[2]{\State \algorithmicif\ #1\ \algorithmicthen\ #2}% \IfThenElse{<if>}{<then>}
\algnewcommand{\IfThenElse}[3]{\State \algorithmicif\ #1\ \algorithmicthen\ #2\ \algorithmicelse\ #3}% \IfThenElse{<if>}{<then>}{<else>}
%\DeclareMathOperator{\card}{card}
%\DeclareMathOperator{\Flat}{Flat}
\renewcommand{\P}{\mathcal P}
\usepackage{listings}
\lstset{basicstyle=\small}
%\usepackage{algorithm} 
%\usepackage{algpseudocode}

@@

@@

\section{BIP architectures, and their encodings into pNets}
\label{section:BIParchitectures}
\label{section:examples}

@@

\section{Running example}
\label{section:FailureTimerMax}
\TODO{1.5 page including figures ?}

@@

 \begin{figure}[t]
   \centering
   \includegraphics[width=\columnwidth]{XFIG/BIPspec-ArchFailureTimerMax-v2}
   \caption{The BIP specification of the FailureTimer architecture}
   \label{schema:ArchFailure:BIP}
 \end{figure}
%
   %% \hfill
%
 \begin{figure}[t]
   \centering
   \includegraphics[width=\columnwidth]{XFIG/FailureTimerPNET}
   \caption{pNet encoding of the FailureTimer architecture}
   \label{schema:ArchFailure:pNet}
 \end{figure}

@@

@@

\newpage
\appendix
%\section{Appendix}
The following appendices are included for the convenience of reviewers.
Our intention is to publish as a research report, before final publication, an
extended version containing all technical detials, proofs, and more examples.

These are organized in th efollowing way:

\begin{enumerate}
\item[A] pNet semantic rules
\item[B] Variable managment
\item[C] Translation to SMTlib input language
\item[D] Well-formness and typing of expressions
\item[E]

\end{enumerate}

@@

% \section{More details on the generation algorithm}

\section{Variable managment}
The variables in each synchronisation vector are considered local: for
a given pNet expression, we must have fresh local variables for each
occurrence of a vector (= each time we instantiate rule
Tr2). Similarly the state variables of each copy of a given pLTS in
the system, must be distinct, and those created for each application
of Tr2 have to be fresh and all distinct. This will be implemented
within the open-automaton generation algorithm, e.g. using name
generation using a global counter as a suffix.
\TODO{add structure of generated names}

\section{Translation to SMTlib input language}
In order to submit satisfiability problems to Z3 (for the predicates
in open transitions), we need to generate SMTlib programs, from the
pNet Algebra presentation and predicates.
More precisely, we need to translate to SMTlib:
\begin{itemize}
  \item the presentation of the action algebra
    (sorts and operators) that is defined for a given language (process calculus, or high level programming language),
    \item for a given pNet, the set of local constants (actions or auxiliary data) that are used in the pLTSs,
    \item for each open transition: the declaration of variables, and
      the predicate (including action expressions).
\end{itemize}

@@

\subsection{Definition of an algebra presentation}
We have a minimal, predefined algebra presentation for all pNets, including
three basic sorts $Bool$, $Action$ and $Int$ and their
operators. Table \ref{Table:predefinedSorts} defines these elements.

@@

\begin{table}\caption{\label{Table:predefinedSorts}\leftline{Algebra Presentation: predefined Sorts
  and Operators}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Other Operators
                \\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
                $\texttt{true},\ \texttt{false}$&
                $\land,\ \lor,\ \neg,\ \implies$
                \\\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{FUN}$ &
                \\\specialrule{0em}{1pt}{1pt}
		Int 				&
                ${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
                $- \texttt{(unary)},\ +,\ -
                \texttt{(binary)},\ \times,\ \div, \texttt{etc.}$
                \\\specialrule{0em}{1pt}{1pt}
                \textsl{for any sort} & & $=,\ \ne$
		\\\hline
	\end{tabular}
\end{table}

\begin{table}\caption{\label{table:BIPalgebra}\leftline{Additional
      operator for the BIP algebra}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Other Operators
                \\\hline\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{FUN\_Action\_Bool}$ &
		\\\hline
	\end{tabular}
\end{table}

@@

\begin{table}\caption{\label{table:BIPalgebra}\leftline{Additional
      sorts and operators for the CCS algebra}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Other Operators
                \\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Channel, Data    			&
                $\emptyset$&
                $\emptyset$
                \\\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{Emit(c,v)},\texttt{Receive(c,v)},\texttt{Tau}$ &
		\\\hline
	\end{tabular}
\end{table}

Then in pNet API these operators or constants can be declared as:
\begin{lstlisting}[basicstyle=\scriptsize\ttfamily, language=java, frame=single]
AlgebraSort Action = new AlgebraSortImpl("Action");
AlgebraSort Channel = new AlgebraSortImpl("Channel");
AlgebraSort Data = new AlgebraSortImpl("Data");
Action.addConstructor("Emit", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Receive", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Tau");
\end{lstlisting}

%\subsection{Constants and variables in pNets and in open transitions}
\subsection{Extended presentation, and Environment}

@@

\section{Well-formness and typing of expressions}

The purpose of this section is to define static semantic notions that
will guarantee that the translation to the SMT language will be
correct, i.e. will not yield errors at runtime. This includes
well-formness (all sorts, operators, variables are defined, and
expressions respect the arity of operators), and typing rules.
%\TODO{Eric: refer in the appendix header}

@@

\section{Map to SMT-LIB language}
The pNet elements, as defined above, can be full translated into
SMT-LIB language, but there are a number of differences in the
structure of the models/langages, so the translation is not trivial.

In this section we shall define separately the translation of the
extended algebra presentation for one pNet (so it will be common for
the study of all OTs of one use-case); and the translation of each
predicate (of an open-transition) of this pNet.

@@

\subsection{Simplification of Predicates}
\subsection{Simplification of Predicates}
\ERIC{: motivated, add refs, and an abstract presentation of stuff
  from pages 19-20. The following is NOT a good example, we should
  rather use a pNet with at least 2 levels, to have real intermediate
variables; here b0, b1, b2 are NOT intermediate variables... and the
predicate simplification is using `some form of ``tautology checking''
rather thanm rewritting.}

>>>>

\input{XFIG/macros.tex}         % Definitions of macros used in XFig figures

@@

\newcommand{\mdash}{---}
\newcommand{\ndash}{--}
\newcommand{\cf}[1][\ ]{cf.#1}
\newcommand{\ie}[1][\ ]{i.e.#1}
\newcommand{\etc}[1][\ ]{etc.#1}
\newcommand{\eg}[1][\ ]{e.g.#1}
\newcommand{\wrt}[1][\ ]{w.r.t.#1}

\newcommand{\goesto}[2][]{\ensuremath{\xrightarrow[#1]{#2}}}
\newcommand{\true}{\ensuremath{\mathit{true}}}

@@

% Macros for the SOS rules and proof trees:
%\newcommand\openrule[2]{\redinfer{#1}{#2}}
\newcommand\openrule[2]{\inferrule*[myfraction=\reddottedrule,center]{#1}{#2}}
%\newcommand\openrule[2]{\inferrule*{#1}{#2}}

%% SB: Moved the macro definitions from XFIG/figtex2eps-preamble.tex
%%     to XFIG/macros.tex and added a com and to input this latter one
%%     into the paper source (above in this file).
%\newcommand\ostate[1]{\triangleleft{\;#1\;}\triangleright}
%\newcommand\ostate[1]{\triangleleft{#1}\triangleright}
\newcommand{\sm}[1]{\mbox{\boldmath\small #1}}
\usepackage{algorithm} 
%\usepackage{algorithmicx} 
\usepackage{algpseudocode}  
\renewcommand{\algorithmicrequire}{\textbf{Input:}} 
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\algnewcommand{\IfThen}[2]{\State \algorithmicif\ #1\ \algorithmicthen\ #2}% \IfThenElse{<if>}{<then>}
\algnewcommand{\IfThenElse}[3]{\State \algorithmicif\ #1\ \algorithmicthen\ #2\ \algorithmicelse\ #3}% \IfThenElse{<if>}{<then>}{<else>}
%\DeclareMathOperator{\card}{card}
%\DeclareMathOperator{\Flat}{Flat}
\renewcommand{\P}{\mathcal P}
\usepackage{listings}
\lstset{basicstyle=\small}
%\usepackage{algorithm} 
%\usepackage{algpseudocode}

@@

\TODO{To be reduced, moving more material to the appendix}

@@

\section{Running example}
\label{section:FailureTimerMax}
\label{section:BIParchitectures}
\label{section:examples}

As a running example we use the Failure Monitor architecture from the
CubETH nanosatellite on-board software
case-study~\cite{CubETH-case-study} realised using BIP.  BIP
(Behaviour-Interaction-Priority)~\cite{bip} is a framework for the
component-based design of concurrent software and systems.  In
particular, the BIP tool-set comprises compilers for generating C/C++
code, executable by linking with one of the dedicated engines, which
implement the BIP operational semantics.  This approach ensures that
any property, shown to hold on a given BIP model, will also hold by
construction on the generated code.

In this context, the notion of \emph{architecture} was proposed
in~\cite{AttieBBJS16-architectures-faoc} as a mechanism for ensuring
correctness by construction during the design of BIP models.
Architecutres can be viewed as operators transforming BIP models.
They formalise design patterns, which enforce global properties
characterising the coordination among the components of the system.
The architecture-based design process in BIP takes as input a set of
components providing basic functionality of the system and a set of
temporal properties that must be enforced in the final system.  For
each property, a corresponding architecture is identified\,\footnote{%
%
  The current approach relies on taxonomies of predefined
  architectures.  Future work may include automatic synthesis of
  architectures from, \eg CTL or a similar logic. 
%
} and applied to the model, thereby potentially introducing additional
coordinator components and modifying the connectors that define
synchronisation patterns among ports of components.
In~\cite{AttieBBJS16-architectures-faoc}, it was shown that
application of architectures is compositional \wrt safety properties,
\ie when several architectures are applied, each enforcing a safety
property, the resulting system satisfies their conjunction.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{XFIG/BIPspec-ArchFailureTimerMax-v2}
  \caption{The BIP specification of the Failure Monitor architecture}
  \label{schema:ArchFailure:BIP}
\end{figure}

Figure~\ref{schema:ArchFailure:BIP} shows a refined version of the
Failure Monitor architecture used in~\cite{CubETH-case-study}.
Contrary to standard BIP models, architectures comprise one or several
\emph{operand} components, whereof only the set of \emph{ports} is
given.  Here, the operand component is $B$ and its interface consists
of the ports $\PortFinish$, $\PortResume$ and $\PortFail$.  The two
\emph{coordinator} components\mdash $\NameCtrl$ and $\NameTimer$\mdash
are standard BIP components insofar as they also have their
\emph{behaviour} specified by finite automata extended with local data
variables.  Transitions of these automata are labelled with the ports
of the corresponding component, Boolean guards and update functions on
local variables.  For instance the loop transition $\tstate{1}
\goesto{\PortTick, \GuardTick[], \TimerUpd[]} \tstate{1}$ in the
$\NameTimer$ component is labeled by the port $\PortTick$, it can be
fired only when the current value of the local variable $t$ is greater
than $0$.  Upon firing, this transition decrements the value of $t$ by
$1$.  When omitted, the default guard (resp. update function) is the
constant predicate $\true$ (resp. the $\mathit{skip}$ operator).  The
constant $\mathrm{Max}$, in $\tstate{0} \goesto{\PortStart,
  \TimerInit[]} \tstate{1}$, is a parameter of the architecture.

\TODO{1.5 page including figures ?}

\TODO{Simon: can you do this in less than one page ? If you prefer,
  merge with ``Running example'' may be easier to describe the syntax}

@@

@@

 \begin{figure}[t]
   \centering
   \includegraphics[width=0.9\columnwidth]{XFIG/FailureTimerPNET}
   \caption{pNet encoding of the FailureTimer architecture}
   \label{schema:ArchFailure:pNet}
 \end{figure}

@@

\subsection{Simplification of Predicates}
\ERIC{: motivated, add refs, and an abstract presentation of stuff
  from pages 19-20. The following is NOT a good example, we should
  rather use a pNet with at least 2 levels, to have real intermediate
variables; here b0, b1, b2 are NOT intermediate variables... and the
predicate simplification is using `some form of ``tautology checking''
rather thanm rewritting.}

@@

\newpage
\appendix
\section{Appendix}
The following appendices are included for the convenience of reviewers.
Our intention is to publish as a research report, before final publication, an
extended version containing all technical detials, proofs, and more examples.

@@

\subsection{Variable managment}
The variables in each synchronisation vector are considered local: for a given pNet expression, we must have fresh local variables for each occurrence of a vector (= each time we instantiate rule Tr2). Similarly the state variables of each copy of a given pLTS in the system, must be distinct, and those created for each application of Tr2 have to be fresh and all distinct. This will be implemented within the open-automaton generation algorithm, e.g. using name generation using a global counter as a suffix.

\subsection{Translation to SMTlib input language}
In order to submit satisfiability problems to Z3 (for the predicates
in open transitions), we need to generate SMTlib programs, from the
pNet Algebra presentation and predicates.
More precisely, we need to translate to SMTlib:
\begin{itemize}
  \item the presentation of the action algebra
    (sorts and operators) that is defined for a given language (process calculus, or high level programming language),
    \item for a given pNet, the set of local constants (actions or auxiliary data) that are used in the pLTSs,
    \item for each open transition: the declaration of variables, and
      the predicate (including action expressions).
\end{itemize}

@@

We have a minimal, predefined algebra presentation for all pNets, including
three basic sorts $Bool$, $Action$ and $Int$ and their
operators. Table \ref{Table:predefinedSorts} defines these elements.

@@

\begin{table}\caption{\label{Table:predefinedSorts}\leftline{Algebra Presentation: predefined Sorts
  and Operators}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Other Operators
                \\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
                $\texttt{true},\ \texttt{false}$&
                $\land,\ \lor,\ \neg,\ \implies$
                \\\specialrule{0em}{1pt}{1pt} 
		Action 			&  $texttt{FUN}$ &
                \\\specialrule{0em}{1pt}{1pt}
		Int 				&
                ${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
                $- \texttt{(unary)},\ +,\ -
                \texttt{(binary)},\ \times,\ \div, \texttt{etc.}$
                \\\specialrule{0em}{1pt}{1pt}
                \textsl{for any sort} & & $=,\ \ne$
		\\\hline
	\end{tabular}
\end{table}

@@

Then in pNet API these operators or constants (seemed as non-parameter operators) can be declared as:
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily, language=java, frame=single]
AlgebraSort Action = new AlgebraSortImpl("Action");
AlgebraSort Channel = new AlgebraSortImpl("Channel");
AlgebraSort Data = new AlgebraSortImpl("Data");
Action.addConstructor("Emit", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Receive", 
	{"Chan_E", "Value_E"}, {Channel, Data});
Action.addConstructor("Tau");
\end{lstlisting}

\paragraph{Expressions}
Once an algebra presentation is defined, we can construct expressions
of the term algebra, using variables and operators:

\begin{definition}
  Let $\mathcal{V}$ be a set of variables. The Term Algebra
  $\Sigma_{\mathcal{P},\mathcal{V}}$ is the set of:
  \begin{itemize}
  \item variables $x \in \mathcal{V}$
  \item terms (or expressions) $op (arg_1, ... , arg_n)$ with $op$ an operator in
  $\mathcal{P}$ of arity $n$, and each $arg_i$ an expression.
  \end{itemize} 
\end{definition}

\paragraph{Constants and variables in pNets and in open transitions}

@@

\paragraph{Well-formed and well-typed expressions}

The purpose of this section is to define static semantic notions that
will guarantee that the translation to the SMT language will be
correct, i.e. will not yield errors at runtime. This includes
well-formness (all sorts, operators, variables are defined, and
expressions respect the arity of operators), and typing rules.
\TODO{Eric: refer in the appendix header}

@@

\subsubsection{Map to SMT-LIB language}
The pNet elements, as defined above, can be full translated into
SMT-LIB language, but there are a number of differences in the
structure of the models/langages, so the translation is not trivial.

In this section we shall define separately the translation of the
(extended) algebra presentation for one pNet (so it will be fixed for
the study of one use-case); and the translation of each predicate in
the context of an open-transition of this pNet.

@@

\subsection{Simplification of Predicates}
\TODO{[Eric]: Motivate, and refs  -- reduce the section here to approx
  half a page (from the next two pages...)}
